---
title: "LMA project - EDA"
author: "Huy Le"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_notebook
---
Commented header  
=========================================

```{r echo = TRUE}
# Course: 
# Title: 
# Purpose: 
# Date: 
# Author: Huy LE
```


Clear environment of variables and packages  
=========================================

```{r include=FALSE}
# A common problem in coding is having unknown dependencies on variables or 
# packages being used in earlier analysis. This happens when you are running 
# code but do not close R or RStudio between sessions. To reduce the probability
# of this happening you can clear your environmen t of variables, functions and
# packages at the beginning of a script. That way you know your code is only
# running on what is within that script.
# Note: you can reorder a long comment with short-cut Ctrl+Shift+/

# Clear environment of variables and functions
rm(list = ls(all = TRUE)) 

# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)

```

Load packages that include different functions  
=========================================


```{r include=FALSE}
# Load libraries
library(tidyverse)
# WILL ONLY USE BASE CODE IN THIS ANALYSIS SO NOT LOAD LIBRARIES

```

Load data and begin new EDA  
=========================================


```{r}
# Load data 
wapop_data <- read.csv("../2.PreparedData/wadata.csv")

# Examine data structure 
head(wapop_data$ID)
str(wapop_data)

```
Prepare data sample for EDA
==========================================
```{r}
# prepare data for EDA
# Create a sample for EDA
#### wapop_data[is.na(wapop_data$SCHL) <- 0] -- commented SG
wapop_data[is.na(wapop_data)] <- 0 # -- Added SG

# labling the variable RAC1P
levels(wapop_data$RAC1P) <- list("1" = "White alone",
                                "2" =  "Black or African American alone",
                               "3" =  "American Indian alone",
                               "4" =  "Alaska Native alone",
                               "5" =  "American Indian and Alaska Native",
                               "6" =  "Asian alone",
                               "7" =  "Native Hawaiian and Other Pacific Islander alone",
                               "8" =  "Some Other Race alone",
                               "9" =  "Two or More Races")



wapop_data <- wapop_data %>% mutate(EDU = ifelse(SCHL > 15, ifelse( SCHL > 19, "HigherEduc", "Associate"), "HighSchool")) %>%
                         select(ID, RAC1P, EDU, SCHL, PERNP )

wapop_filter <- wapop_data %>% filter(RAC1P != 9, SCHL != 0) 

set.seed(2)
wapop_sample <- wapop_filter[sample(nrow(wapop_filter),500),]


str(wapop_sample)

summary(wapop_sample)
```


1. Univariate non-graphical -- Categorical -- before excluded
=========================================
##Step 1

```{r}
Race_tab <- rbind(table(wapop_data$RAC1P),
                    prop.table(table(wapop_data$RAC1P)),
                    prop.table(table(wapop_data$RAC1P))*100)
row.names(Race_tab) <- c("Count", "Proportion", "Percent")
round(Race_tab, 2)

```

+ 78.41% of Pop is White alone (na.rm =TRUE)
+ 5.12% is two or more than two race 
    * as we discussed, we will remove those observation with race == 9
    

##Step 2

```{r}
shcl_tab <- rbind(table(wapop_data$SCHL),
                    prop.table(table(wapop_data$SCHL)),
                    prop.table(table(wapop_data$SCHL))*100)
row.names(shcl_tab) <- c("Count", "Proportion", "Percent")
round(shcl_tab, 2)

# number observation have SCHL = NA
sum(is.na(wapop_data$SCHL))
sum(is.na(wapop_data$SCHL))/nrow(wapop_data)*100
#
# nrow(wapop_data[wapop_data$SCHL <= 15,])
# nrow(wapop_data[wapop_data$SCHL <= 15,])/nrow(wapop_data)*100
educ_tab <-  rbind(table(wapop_data$EDU),
                    prop.table(table(wapop_data$EDU)),
                    prop.table(table(wapop_data$EDU))*100)
row.names(educ_tab) <- c("Count", "Proportion", "Percent")
round(educ_tab, 2)
```

- 2317 (3,2%) person are under 3 years old (Education == N.A) => will remove  when doing futher analysis
- 17% of population have Bachelor' Degree (21)
- 15% of population obtained high school diploma (16)
- 17730 (na.rm =TRUE) (25.30%) people are in <= high school

##Step 3

```{r}
summary(wapop_data$PERNP)

# Na observation in earning
nrow(wapop_data[wapop_data$PERNP == 0,])

```
================================================================
Min.      1st Qu.   Median      Mean    3rd Qu.    Max.     NA's 
  -5100       0     14000       33897   48000     850000    13270 
=================================================================
- 2317 < 3 yrs, but 13270 obs NA in earning and 34241 have 0 earnings
    + It could be because 20047 people are in or lower highschool
    
- Should change the unit of earnings  from $ -> 1000 $ for visualization.    

    
2. Univariate graphical code 
=========================================
in this section we will user wapop_sample, for EAD, because the population is too big

    2.1 Education Attainment

```{r}

# graphics barplots to see count distribution of factor levels
p <- ggplot( data = wapop_sample, mapping = aes( x = EDU))

p + geom_bar(stat = "count") +
      ggtitle("WA population Education attainment in sample 500obs") +
      geom_text(stat = "count", 
                aes(label = ..count..), 
                vjust = -0.2, 
                position = position_dodge(0.9))+
      labs(x = "Education attainment", y = "Frequence")

```

    2.2 Race

```{r}
p <- ggplot( data = wapop_sample, mapping = aes( x = RAC1P, fill = factor(RAC1P)))

p + geom_bar(stat = "count") +
      ggtitle("WA population Races and Ethnicities in sample 500obs") +
      geom_text(stat = "count", 
                aes(label = ..count..), 
                vjust = -0.2, 
                position = position_dodge(0.9))+
      labs(x = "Races and Ethnicities", y = "Frequence", fill = "Race")

```


3. Univariate non-graphical -- Quantitative
=========================================

```{r}
# Get descrptive stats 
summary(wapop_data)
summary(wapop_sample)
```

+ All three numerical and 1 integer variables have divergent mean and median so likely skewed distributions  

4. Univariate graphical -- Quantitative
=========================================
Histograms 
```{r}

```


```{r}
p <- ggplot( data = wapop_sample, mapping = aes( x = RAC1P, y = PERNP/1000, colour = EDU))

p + geom_point(stat = "identity") +
      ggtitle("WA population Earnings in Races") +
      labs(x = "Race", y = "Earnings")
# 
#      geom_text(stat = "identity", 
#                 aes(label = ..count..)
#                ) +

hist(wapop_sample$PERNP/1000, xlim = c(min(wapop_sample$PERNP/1000, na.rm = TRUE), max(wapop_sample$PERNP/1000, na.rm = TRUE))) #unit 1000 USD

par(mfrow = c(1,2))
plot(wapop_sample$RAC1P, wapop_sample$PERNP/1000 ,pch = 16)
plot(wapop_sample$SCHL, wapop_sample$PERNP/1000 ,pch = 16)

# levels(wapop_sample$SCHL) <- edu_title
# nlevels(wapop_sample$SCHL)
# sort(unique(wapop_sample$SCHL))
# levels(factor(wapop_sample$SCHL))

```
- High income places almost at "White Alone"" and people have at least Bachelor's Degree.


## Boxplots of quantitative data 

```{r}
# Set grid pattern for graph arrangement
#par(mfrow = c(2, 3))
#boxplot(office$Year, main = "Year")
#boxplot(office$Sales, main = "Sales")

```


5. Multivariate non-graphical -- Categorical
=========================================
Use cross-tabs for counts/proportion/percent  

+ Cross-tabs are for more than one categorical variable  

+ Create cross-tab with margins

```{r}
# Make a cross-tab with margins
addmargins(xtabs(~ EDU + RAC1P,  data = wapop_sample))

```

+ But proportions are easier to understand so make a proportions table

```{r}
# Make a cross-tab with margins
addmargins(round(prop.table(xtabs(~ EDU + RAC1P,  data = wapop_sample)),3))

```



```{r}
# Make a cross-tab with margins
#addmargins(prop.table(xtabs(~ Region + Sub.Category,  data = office)))

```


6. Multivariate graphical code -- Categorical
=========================================

+ Tile graph  

```{r}
#Correlation Race - Education

p <- ggplot( data = wapop_sample, mapping = aes( x = RAC1P, fill= EDU))

p + geom_bar(stat = "count", position = "dodge") +
    ggtitle("WA population Races - Education") +
    geom_text(stat = "count",
                aes(label = ..count..),
                vjust = -0.2, 
                position = position_dodge(0.9)) +
    labs(x = "Race", y = "Education")

```


7. Multivariate non-graphical -- Quantitative
=========================================
The standard measure between quantitative variables is correlation  

```{r}
# Find correlation of quantiative variables
str(wapop_sample)
plot(wapop_sample$RAC1P, wapop_sample$SCHL)


```


8. Multivariate graphical code -- Quantitative
=========================================

```{r}
# Correlation plot
#plot(office[, 5:8])

```

+ Note: can see correlations and individual variable distributions  

```{r}
# I prefer the correlation plot from the psych package
#library(psych)
#pairs.panels(office[, 5:8])

```

+ See an interesting relationship between profit and the other quantitative variables


## There are many options when using quantitative variables

Usually some form of a scatter plot

+ Can use size, shape, color to identify additional variables  
+   


```{r}
# Begin with standard scatter plot
#plot(office$Discount, office$Profit)

```

+ This is just a larger version of the plot above  
    - Use if you want to examine a pattern more closely  
+ Profits decrease as discounts increase  


## Can add other variables with color, shape, size, etc


```{r}
# Differentiate region with color, not they are alphabetical
#plot(office$Discount, office$Profit, col = office$Region)
#legend(x = "topright",  
#       legend = levels(office$Region), 
#       col = c(1:nlevels(office$Region)), 
#       pch = 1) 

```


9. Multivariate non-graphical -- Categorical and Quantitative
=========================================

```{r}
# Form table of total profit by by region and sub.category
# tapply function is like other apply functions but does a function [sum in this case] over an indicie [Region in this case]
# apply functions are generally
#profit_sum <- round(rbind(
#  tapply(office[office$Region == "Central",]$Profit, office[office$Region == "Central",]$Sub.Category, sum),
#  tapply(office[office$Region == "East",]$Profit, office[office$Region == "East",]$Sub.Category, sum),
#  tapply(office[office$Region == "South",]$Profit, office[office$Region == "South",]$Sub.Category, sum),
#  tapply(office[office$Region == "West",]$Profit, office[office$Region == "West",]$Sub.Category, sum)
#  ),
#  0)
#row.names(profit_sum) <- c("Central", "East", "South", "West")
#profit_sum


```


```{r}
# Form table of medians since profit was skewed



```



10. Multivariate graphical -- Categorical and Quantitative
=========================================
Many options available  

+ First do a basic barplot    
    - Will try many others when use ggplot2  
+ Again, these are exploration graphs  
    - Not presentation graphs  
    
```{r}
# Doing a subset of the table to just show variables of interest
#keep_cat <- c("Accessories", "Binders", "Copiers", "Machines", "Paper", "Phones", "Tables")
#small_tab <- profit_sum[, keep_cat]

# Note: had to use t() to traspose the table to get it to print with profit in the bars and Sub.Cat on x-axis
#par(mfrow = c(1, 1))
#barplot(small_tab, 
#         beside = TRUE, 
#         horiz = TRUE, 
#         las = 2, 
#         col = c(1:4),
#         main = "Stop or reduce selling machines.\nProducts to increase varies by region.",
#         xlab = "Total profit by region and subcategory",
#         adj = 0)
# axis(1, at=axTicks(1), labels=sprintf("$%s", axTicks(1)), las = 2) # add $ to x-axis tics
# legend(x = "topright",  
#        legend = levels(office$Region), 
#        col = c(1:4),
#        pch = 15)

```


Now I am going to make a copy of the graph that I can call from another RMarkedown file  

+ Could also save as jpeg or pdf  

```{r}
#  
# pdf(NULL)
# dev.control(displaylist = "enable")
# 
# par(mfrow = c(1, 1))
# barplot(small_tab, 
#         beside = TRUE, 
#         horiz = TRUE, 
#         las = 2, 
#         col = c(1:4),
#         main = "Stop or reduce selling machines.\nProducts to increase varies by region.",
#         xlab = "Total profit by region and subcategory",
#         adj = 0)
# axis(1, at=axTicks(1), labels=sprintf("$%s", axTicks(1)), las = 2) # add $ to x-axis tics
# legend(x = "topright",  
#        legend = levels(office$Region), 
#        col = c(1:4),
#        pch = 15)
# 
# profit_plot <- recordPlot()
# 
# invisible(dev.off())
# 
# profit_plot

```



Summary
=========================================


```{r}
# Save the rds file so I can reuse anything from this file in another file
# save.image("office_TA.RData")

# recover using load()
# load("office_TA.RData")
```

