---
title: "R Notebook"
output: html_notebook
---


```{r}

```

Clear environment of variables and packages  
=========================================

```{r}

# Clear environment of variables and functions
rm(list = ls(all = TRUE)) 

# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)

```

Load packages that include different functions  
=========================================


```{r }
# Load libraries
library(tidyverse)
library(fastDummies)
library(lmtest)
library(sandwich)
library(GGally)
library(olsrr)
library(robustbase)

```

Load data and transform 
=========================================

```{r}
# Load data 
waraw <- read.csv("../2.PreparedData/wadata3.csv")

```



```{r}
# labling the variable RAC1P
Race_tab <- data.frame("RAC1P" = 1:9,
                        "Races"= c("White",
                              "Black",
                              "American Indian",
                              "Alaska Native",
                              "American Indian and Alaska Tribes",
                              "Asian",
                              "Islander",
                              "Others",
                              "Two or More Races"))


Gender_tab <- data.frame("SEX" = 1:2, "Gender" = c("Male", "Femeal"))

Mar_tab <- data.frame("MAR" = 1:5,
                      "Marital" = c("Married","Widowed","Divorced","Separated","Never Married"))

# clean data
wa <- waraw %>% 
      select(AGEP, RAC1P, SCHL, MAR, SEX, PERNP, SCIENGP, SCIENGRLP, LANX, WKHP) %>%
      # remove all obser under 14 years old, NA value  in education & earnings, Earning < 0
      filter(AGEP >= 14, !is.na(waraw$PERNP), !is.na(waraw$SCHL), RAC1P != 9, PERNP >= 300) %>%
      mutate(EDU = ifelse(SCHL <= 9, "Primary_School", 
                          ifelse( SCHL < 16, "Under_Highschool", 
                                  ifelse( SCHL == 16, "Highschool",
                                          ifelse( SCHL <= 19 ,"Some_College",
                                                 ifelse( SCHL == 20 ,"Associate",
                                                         ifelse( SCHL == 21 ,"Bachelor",
                                                                 ifelse( SCHL == 22 ,"Master",
                                                                         ifelse( SCHL == 23 ,"Professional", "Docterate")))))))),
             IsFullTime = ifelse(WKHP > 34 & !is.na(WKHP), 1, 0),
             IsScienEngr = ifelse(SCIENGP == 1 & !is.na(SCIENGP), TRUE, FALSE),
             OnlyEnglish = ifelse(LANX == 2, TRUE, FALSE))%>%
      left_join(Race_tab, by = "RAC1P") %>%
      left_join(Gender_tab, by = "SEX") %>%
      left_join(Mar_tab, by = "MAR")

head(wa)
wa$EDU = factor(wa$EDU, levels=c('Primary_School', 'Under_Highschool', 'Highschool','Some_College', 'Associate', 'Bachelor', 'Master', 'Professional', 'Docterate')) 
wa$IsFullTime <- as.factor(wa$IsFullTime)
wa$IsScienEngr <- as.factor(wa$IsScienEngr)
wa$OnlyEnglish <- as.factor(wa$OnlyEnglish)



summary(wa)
nrow(is.na(wa$EDU))
```

##1.	In the state of Washington, how do earnings vary by educational attainment?
### First model Earning ~ edu + AGEP + AGEP^2 +others
```{r}
#Prepare data
wadata_mr <- wa %>% 
              #filter(IsFullTime == 1) %>%
              select(PERNP, EDU, AGEP, Marital, Gender, IsScienEngr ) %>%
              fastDummies::dummy_cols(remove_first_dummy = TRUE) %>%
              select( -EDU, -IsScienEngr, -Marital, -Gender)

summary(wadata_mr)
# SLR Earning ~ education age.
Ear.edu.model <- lm(data = wadata_mr, PERNP ~ . + I(AGEP*AGEP))

summary(Ear.edu.model)

```

### Testing Collinearity
```{r}
# Collinearity diag
vcovHC(Ear.edu.model, type = "HC")
#Generate the Robust standard errors and print them on screen 
sandwich_se <- sqrt(diag(vcovHC(Ear.edu.model, type = "HC")))
sandwich_se

```

### BP test

```{r}
bptest(Ear.edu.model)
bptestEarn.model = lm(data = wadata_mr, residuals(Ear.edu.model)*residuals(Ear.edu.model) ~ . + I(AGEP*AGEP))
summary(bptestEarn.model)
```
studentized Breusch-Pagan test

data:  Ear.edu.model
BP = 1254.5, df = 16, p-value < 2.2e-16

p-value too small => reject H null => heteroskedaticity exists
 => estimate logarith model

### Log model
```{r}
# Logarith Earning ~ education age.
Eaning.Edu.Log.model <- lm(data = wadata_mr, log(PERNP) ~ . + I(AGEP*AGEP))

summary(Eaning.Edu.Log.model)
```

### BP test log(earning) ~...
```{r}
bptest(Eaning.Edu.Log.model)
bptest.logEarn.model = lm(data = wadata_mr, residuals(Eaning.Edu.Log.model)*residuals(Eaning.Edu.Log.model) ~ . + I(AGEP*AGEP))
summary(bptest.logEarn.model)
```

	studentized Breusch-Pagan test

data:  Eaning.Edu.Log.model
BP = 662.04, df = 16, p-value < 2.2e-16

# Robust linear model
```{r}
# robust standard error
Robus.Se <- sqrt(diag(vcovHC(Eaning.Edu.Log.model, type = "HC")))
coeftest(Eaning.Edu.Log.model, vcov = vcovHC(Eaning.Edu.Log.model,type = "HC"))

# robust standard error linear model
robust.log.earn <- lmrob(log(PERNP) ~ . + I(AGEP*AGEP),  data = wadata_mr)

summary(robust.log.earn)
```


### Premium Earning(higher education) ~ Race

```{r}
wadata_q2 <- wa %>% 
              filter(SCHL >= 17) %>%
              select(PERNP, Races, AGEP, Marital, Gender, IsScienEngr) %>%
              fastDummies::dummy_cols(remove_first_dummy = TRUE) %>%
              select( - c(Races, Gender, Marital, IsScienEngr))

# MLR Earning ~ education age.
Premium.Earning.model <- lm(data = wadata_q2, PERNP ~. + I(AGEP^2))

summary(Premium.Earning.model)
```

### Testing Collinearity
```{r}
# Collinearity diag
vcovHC(Premium.Earning.model, type = "HC")
#Generate the Robust standard errors and print them on screen 
premium_se <- sqrt(diag(vcovHC(Premium.Earning.model, type = "HC")))
premium_se

```

### BP test

```{r}
# BP test for Premium model
bptest(Premium.Earning.model)
bptestPremium.model = lm(data = wadata_mr, residuals(Premium.Earning.model)*residuals(Premium.Earning.model) ~ . + I(AGEP*AGEP))
summary(bptestPremium.model)
```
	studentized Breusch-Pagan test

data:  Premium.Earning.model
BP = 615.26, df = 15, p-value < 2.2e-16


p-value too small => reject H null => heteroskedaticity exists
 => estimate logarith model

### Log model
```{r}
# Logarith Earning ~ education age.
Premium.Log.model <- lm(data = wadata_q2, log(PERNP) ~ . + I(AGEP*AGEP))

summary(Premium.Log.model)
```

### BP test log(earning) ~...
```{r}
# BP test for log( earning ) premium model
bptest(Premium.Log.model)
bptest.logPre.model = lm(data = wadata_q2, residuals(Premium.Log.model)*residuals(Premium.Log.model) ~ . + I(AGEP*AGEP))
summary(bptest.logPre.model)
```

	studentized Breusch-Pagan test

data:  Premium.Log.model
BP = 431.45, df = 15, p-value < 2.2e-16

# Robust linear model
```{r}
Robus.Pre.Se <- sqrt(diag(vcovHC(Eaning.Edu.Log.model, type = "HC")))
coeftest(Premium.Log.model, vcov = vcovHC(Premium.Log.model,type = "HC"))

#linear model with robust standard error
#robust.logPre.model <- lmrob(log(PERNP) ~ . + I(AGEP*AGEP),  data = wadata_q2)

summary(robust.logPre.model)
```

