---
title: "LMA project - EDA"
author: 'Group 4: Han Li - Huy Le - Sourabh Gupta'
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
---

Commented header  
=========================================

```{r echo = TRUE}
# Course: ECON 5300
# Title: 2nd EDA
# Purpose: Preliminary Econometric Estimates 
# Date: Jan 29
# Author: Huy LE
```


Clear environment of variables and packages  
=========================================

```{r}

# Clear environment of variables and functions
rm(list = ls(all = TRUE)) 

# Clear environmet of packages
if(is.null(sessionInfo()$otherPkgs) == FALSE)lapply(paste("package:", names(sessionInfo()$otherPkgs), sep=""), detach, character.only = TRUE, unload = TRUE)

```

Load packages that include different functions  
=========================================


```{r }
# Load libraries
library(tidyverse)
library(fastDummies)

```

Load data and transform 
=========================================


```{r}
# Load data 
waraw <- read.csv("../2.PreparedData/wadata3.csv")


# Examine data structure 
str(waraw)

```


```{r}
# labling the variable RAC1P
Race_tab <- data.frame("RAC1P" = 1:9,
                        "Races"= c("White alone",
                              "Black or African American alone",
                              "American Indian alone",
                              "Alaska Native alone",
                              "American Indian and Alaska Native",
                              "Asian alone",
                              "Native Hawaiian",
                              "Some Other Race alone",
                              "Two or More Races"))


Gender_tab <- data.frame("SEX" = 1:2, "Gender" = c("Male", "Femeal"))

Mar_tab <- data.frame("MAR" = 1:5,
                      "Marital" = c("Married","Widowed","Divorced","Separated","Never Married"))

# clean data
wa <- waraw %>% 
      select(AGEP, RAC1P, SCHL, MAR, SEX, WKHP, PERNP) %>%
      # remove all obser under 14 years old, NA value  in education & earnings, Earning < 0
      filter(AGEP >= 14, !is.na(waraw$PERNP),!is.na(waraw$SCHL), RAC1P != 9, PERNP >= 300) %>%
      mutate(EDU = ifelse(SCHL <= 9, "Under_Highschool", 
                          ifelse( SCHL <= 16, "Highschool", 
                                  ifelse( SCHL <= 19 ,"Some_College",
                                         ifelse( SCHL == 20 ,"Associate",
                                                 ifelse( SCHL == 21 ,"Bachelor",
                                                         ifelse( SCHL == 22 ,"Master",
                                                                 ifelse( SCHL == 23 ,"Professional", "Docterate"))))))),
             IsFullTime = ifelse(WKHP > 34 & !is.na(WKHP), 1, 0))%>%
      left_join(Race_tab, by = "RAC1P") %>%
      left_join(Gender_tab, by = "SEX") %>%
      left_join(Mar_tab, by = "MAR")

head(wa)
wa$EDU <- as.factor(wa$EDU)
wa$IsFullTime <- as.factor(wa$IsFullTime)


summary(wa)
```


1. Univariate non-graphical -- Categorical
=========================================
##Earning by education attainment

```{r}
wa$EDU = factor(wa$EDU, levels=c('Under_Highschool', 'Highschool','Some_College', 'Associate', 'Bachelor', 'Professional',
                                 'Master', 'Docterate')) 
earn_educ_tab <- wa %>%
                select(EDU, PERNP) %>%
                group_by(EDU) %>%
                summarise(MinEarn = min(PERNP), MeanEarn = mean(PERNP), MedianEarn = median(PERNP), Max = max(PERNP))
earn_educ_tab


```

##Earning vary by race

```{r}
earn_race_tab <- wa %>% 
                select(Races, PERNP) %>%
                group_by(Races) %>%
                summarise(MinEarn = min(PERNP), MeanEarn = mean(PERNP), MedianEarn = median(PERNP), Max = max(PERNP))
earn_race_tab
```

##Educations vary by Races


```{r}
Edu_race_tab <- wa %>% 
                select(Races, EDU) %>%
                group_by(Races, EDU) %>%
                summarise(Count = n()) %>%
                spread(EDU, Count)
Edu_race_tab


```


2. Univariate graphical code -- Categorical
=========================================
Categorical data - bar plot of counts


```{r}
#
wa$EDU = factor(wa$EDU, levels=c('Under_Highschool', 'Highschool','Some_College', 'Associate', 'Bachelor', 'Professional',
                                 'Master', 'Docterate')) 
Race_edu_barplot <- wa %>% 
                select(RAC1P, EDU) %>%
                group_by(RAC1P, EDU) %>%
                summarise(Count = n()) %>%
                ggplot(mapping = aes(x = RAC1P, y = Count/nrow(wa)*100, fill = EDU)) +
                         geom_bar(stat = "identity", position = "dodge")
                       
Race_edu_barplot             
```

```{r}

ggplot(data = wa, mapping = aes( x = AGEP, y = PERNP/1000)) +
  geom_point(stat = "identity") +
  ggtitle("WA Residents' Earnings Distribution by ages") +
  xlab("Age") +
  ylab("Earnings (1000$)") +
  geom_smooth(method = lm)
```

### Earnings - RACE -EDU
```{r}
# 
# ggplot(data = wa, mapping = aes( x = PERNP/1000, fill = EDU)) +
#   geom_histogram(stat = "count") +
#   ggtitle("WA Residents' Earnings by age") +
#   xlab("Earnings") +
#   #scale_x_discrete(labels =c(levels(wa$RAC1P))) +
#   ylab("count") 
#   

```

### Earnings BoxPlot
```{r}
wa$EDU = factor(wa$EDU, levels=c('Under_Highschool', 'Highschool','Some_College', 'Associate', 'Bachelor', 'Professional',
                                 'Master', 'Docterate')) 
wa %>% ggplot(mapping = aes(x = EDU, y= PERNP/1000)) +
      geom_boxplot() +
      ggtitle("Earnings Distribution") +
      ylab("Total Earnings (1000$)") +
      xlab("Education")

                

```

3. Model
=========================================

###3.1 How do earnings vary by educational attainment? 
#### Model 1-1
```{r}
#Prepare data
wadata_slr <- wa %>% 
              #filter(IsFullTime == 1) %>%
              select(PERNP, EDU, IsFullTime, AGEP) %>%
              fastDummies::dummy_cols(remove_first_dummy = TRUE) %>%
              select( -EDU, -IsFullTime)
# SLR Earning ~ education age.
wa_model <- lm(data = wadata_slr, PERNP ~ .)

summary(wa_model)

```

#### Testing Constant variance model 1

```{r}
#Conduct BPG Test
library(lmtest)
bptest(wa_model)
bptestequation = lm( data=wadata_slr, residuals(wa_model)*residuals(wa_model) ~. - PERNP)
summary(bptestequation) #note: BP = n*RSquared of model with squared residuals as dependent variable 

```
data:  wa_model
BP = 1096.1, df = 9, p-value < 2.2e-16
p-value < 2.2e-16: reject H0 => Heteroskedasticity


### revise model with AGEP^2
#### Model 1-2
```{r}
wa_model_rvs <- lm(data = wadata_slr, PERNP ~ . + I(AGEP^2))

summary(wa_model_rvs)

```

### Testing Homoskedasticity
```{r}
earn_edu_pre <- data.frame("EarnPre"= predict(wa_model_rvs), "Residual" = residuals(wa_model_rvs), "stan.Res" = rstandard(wa_model_rvs), "Fitted"=wadata_slr$PERNP )

#Residual vs fitted polot.
earn_edu_pre %>% ggplot(aes(x = EarnPre, y = Residual)) + geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}

#Conduct BPG Test
bptest(wa_model_rvs)
bptestequation1.2 = lm( data = wadata_slr, residuals(wa_model_rvs)*residuals(wa_model_rvs) ~. + I(AGEP^2) -PERNP)
summary(bptestequation1.2) #note: BP = n*RSquared of model with squared residuals as dependent variable 

library(sandwich)


vcovHC(bptestequation1.2, type = "HC")
#Generate the Robust standard errors and print them on screen 
sandwich_se <- diag(vcovHC(bptestequation1.2, type = "HC"))^0.5
sandwich_se


```
studentized Breusch-Pagan test

data:  wa_model_rvs
BP = 1065.9, df = 10, p-value < 2.2e-16


### estimate log(Earning) model

```{r}
logEarning.model <- lm( data = wadata_slr, log(PERNP)~ . +I(AGEP^2))

summary(logEarning.model)
```

#### Heteroskedasticiy Log equation
```{r}
#Conduct BPG Test
bptest(logEarning.model)
bptestequation1.3 = lm( data = wadata_slr, residuals(logEarning.model)*residuals(logEarning.model) ~. + I(AGEP^2) -PERNP)
summary(bptestequation1.3) #note: BP = n*RSquared of model with squared residuals as dependent variable 
```
studentized Breusch-Pagan test

data:  logEarning.model
BP = 1615.3, df = 10, p-value < 2.2e-16

###3.2 Does the premium for higher education vary by race and ethnicity?

```{r}
wadata_2nd <- wa %>% 
              filter(SCHL >= 17) %>%
              select(PERNP, EDU, IsFullTime, Races, AGEP) %>%
              fastDummies::dummy_cols(remove_first_dummy = TRUE) %>%
              select( - c(Races, EDU))

# MLR Earning ~ education age.
wa_model_q2 <- lm(data = wadata_2nd, PERNP ~. + I(AGEP^2))

summary(wa_model_q2)

```


4.
=========================================
Histograms in base R graphics

```{r}



```


Summary
=========================================


```{r}
# Save the rds file so I can reuse anything from this file in another file
# save.image("office_TA.RData")

# recover using load()
# load("office_TA.RData")
```

